---
title: "Assignment 6. Advanced data visualization"
author: "TBD"
date: "TBD"
output: html_document
---


# Getting started 

Load any packages you will need, then read in the clean penguin data

```{r}

```

# 1.  
Create a scatter plot with regression lines showing how culmen depth varies by body mass, with each species shown in a different color. 
```{r}

```

# 2. 
Create a histogram of culmen depth and overlay a density plot in blue.
```{r}

```

# 3. 
Create a faceted histogram of culmen depth, with a plot for each species. 
```{r}

```

# 4.
Summarize the penguin data and calculate the following statistics on culmen depth for each species and sex: 
    - sample mean 
    - sample standard deviation 
    - sample size 
    - standard error of the mean 
    - upper and lower bounds of a 95% confidence interval on the mean 

```{r}

```

# 5. 
Create a scatter plot showing how mean culmen depth and its confidence interval varies by sex and species. (**Hint**: choose one categorical variable to be your independent variable and display the other by color or shape).
```{r}

```

# 6.
Choose one plot from 1-5 and: 
    
  a) Make the background white without any gridlines  
  b) Make the x and y labels look professional, including units if necessary
  c) Add an informative title
  d) Write a figure legend that completely describes what the figure shows. (You can add this below the R code chunk.)
```{r}

```


# Final steps 

Make sure your name and date are correct in the header. Run spellcheck, then knit this document. Email the knitted HTML to paul.lukacs@umontana.edu.