---
title: "Assignment 5 data visualization"
author: "TBD"
date: "TBD"
output: html_document
---

# 1
Load the clean penguin data. 

```{r}

```

# 2
Create a scatter plot with culmen length as the independent variable, culmen depth as the response variable, and Species showing in different colors.

```{r}

```

# 3
Create a scatter plot with culmen length as the independent variable, culmen depth as the response variable, and every point displayed in the color "green".

```{r}

```

# 4
Create a histogram of flipper length. 

```{r}

```

# 5
Create a boxplot to show how body mass differs by Sex.  

```{r}

```

# 6
Select one of the three previous plots and: 
  a) Change the plot's theme to `theme_classic()`
  b) Change the x and y axis labels to look more professional.
  c) Add an informative plot title (hint: use `ggtitle()`)
  
```{r}

```

# Final steps 

Make sure your name and date are correct in the header. Run spellcheck, then knit this document. Email the knitted HTML to paul.lukacs@umontana.edu.