---
title: "04 Advanced data visualization"
author: "WILD 541"
date: "10/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning objectives
- Add additional layers to a single plot
- Display confidence intervals on a scatter plot
- Visualize a distribution with a density plot
- Create multiple plots using facets

# Picking back up 

Load packages and data
```{r, message = F}
library(tidyverse)
library(here)

penguin <- read_csv(here("data/penguins_clean.csv"))
```

We will start with a scatter plot
```{r, message = F}
ggplot(data = penguin,
       mapping = aes(x = body_mass_g, 
                     y = flipper_length_mm)) + 
  geom_point() 
```

# Adding layers

Whenever you want to add more things to a graph, use `+`. For example, let's add a regression line. 

```{r, message = F}
ggplot(data = penguin,
       mapping = aes(x = body_mass_g, 
                     y = flipper_length_mm)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

When you add more layers to a plot, they will inherit the `aes()` you defined for the whole plot. For example, if we include `Species` in the `aes()` mappings, ggplot automatically builds one regression line for each group and it applies the appropriate color. 

```{r, message = F}
ggplot(data = penguin,
       mapping = aes(x = body_mass_g, 
                     y = flipper_length_mm,
                     color = Species)) + 
  geom_point() + 
  geom_smooth(method = "lm")
```

# Density plot

Instead of a histogram, sometimes we want to display a smooth curve that shows the distribution. This is called a density plot.
```{r, eval = F}
ggplot(penguin, 
       aes(x = body_mass_g)) + 
  geom_histogram(, bins = 10)

ggplot(penguin, 
       aes(x = body_mass_g)) + 
  geom_density()
```

```{r, echo = F}
h <- ggplot(penguin, 
       aes(x = body_mass_g)) + 
  geom_histogram()

d <- ggplot(penguin, 
       aes(x = body_mass_g)) + 
  geom_density()

ggpubr::ggarrange(h, d)
```

To plot the density curve on top of the histogram, we have to change the scale of the y-axis on the histogram from a count to density. (**Note**: this trick comes from [Googling](http://www.sthda.com/english/wiki/ggplot2-density-plot-quick-start-guide-r-software-and-data-visualization) and is not something you would be expected to figure out on your own!)

```{r, message = F}
ggplot(penguin, 
       aes(x = body_mass_g, y = ..density..)) + 
  geom_histogram() + 
  geom_density()
```

# Changing how layers are visualized 

When you add multiple layers to a ggplot, you can change how each of them is visualized by adding arguments inside the GEOM_FUNCTION(). For example, to make your histogram transparent and lump data into fewer bins, you can add `alpha` and `bins` arguments to `geom_histogram()`. To make your density curve red, you can add a `color` argument to `geom_density()`. All the possible arguments are on the Data Visualization cheat sheet.

```{r, message = F}
ggplot(penguin, 
       aes(x = body_mass_g, y = ..density..)) + 
  geom_histogram(alpha = 0.5, bins = 10) + 
  geom_density(color = "red")
```

# Error bars 

To add error bars to a scatterplot, we will first need to summarize the data using **`dplyr`**. Let's visualize mean flipper length of each species and a 95% confidence interval on the mean. (**Stats practice**: Explain what a 95% confidence interval of a mean represents)

```{r}
summ <- penguin %>% 
  group_by(Species) %>% 
  summarize(mean = mean(flipper_length_mm), # Sample mean
            sd = sd(flipper_length_mm), # Sample standard deviation
            n = n() # Sample size
            ) %>% 
  mutate(se = sd/sqrt(n), # standard error of the mean
         lcl = mean - se*1.96, # lower CI bound
         ucl = mean + se*1.96) # upper CI bound

summ
```

To show the errorbars we will need to define `aes()` that are specific to `geom_errorbar()` (`ymin` and `ymax`). You can put these in the global `aes()` we've defined already, but more frequently you will see them in `geom_errorbar()` to make it clear what they're for. 

```{r}
ggplot(summ, 
       aes(x = Species, y = mean)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = lcl, ymax = ucl))
```

Notice the default width is not very attractive. To change the way the errorbars are displayed, add arguments to `geom_errorbar()` *outside* of `aes()`. Again, arguments are found by Googling or looking at the Data Visualization cheat sheet. 
```{r}
ggplot(summ, 
       aes(x = Species, y = mean)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = lcl, ymax = ucl), width = 0.1)
```

*Additional ideas*: if you want to make the x-axis labels not overlap, try Googling "ggplot word wrap x axis labels"

# Facets 

To visualize more factors in your data, you may want to split your plot into multiple plots based on a variable in your data. 

### Facet wrap 

The first way to do this is with `facet_wrap()`. This will create multiple plots and "wrap" them to the next row if you run out of space. 

When using faceting, you will need to take an extra step to make sure **`ggplot`** knows that you are specifying a variable name. (*Why? No one knows*). The syntax is `facet_wrap(vars(<VARIABLE-NAME>))`

```{r}
ggplot(data = penguin,
       mapping = aes(x = body_mass_g, 
                     y = flipper_length_mm, 
                     color = Species)) + 
  geom_point() +
  facet_wrap(vars(Sex))
```


### Facet grid 

Another way to create multiple plots is with `facet_grid()`. You will define variables that create the rows and columns of a grid. 

```{r}
ggplot(data = penguin,
       mapping = aes(x = body_mass_g, 
                     y = flipper_length_mm)) + 
  geom_point() +
  facet_grid(rows = vars(Sex), 
             cols = vars(Species))
```


