---
title: "Assignment 3"
author: "WILD 541"
date: "TBD"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. 
Update the header with your name and today's date. 

# 2. 
Read in penguins_clean.csv file and assign it to an object name
```{r}

```

# 3. 
Create a new object that retains only the following columns: individualID, Species, Stage, culmen_length_mm, culmen_depth_mm, and body_mass_g
```{r}

```

# 4. 
Use the object made in #3 and create a new object keeping only the rows where culmen length is greater than or equal to 48 mm. 
```{r}
```

# 5.
Create a new object from the object created in #4 and add a new column that contains culmen length in inches (hint: to go from mm to in, divide by 25.4)
```{r}

```

# 6. 
String together steps 2-5 so you get the same result as #5 without making any intermediate objects. 
```{r}

```

# 7. 
Run spell check on this document (`Edit > Check Spelling`), then render this document using the **Knit** button. Email the resulting HTML file to paul.lukacs@umontana.edu
