---
title: "Future notes"
author: "Anna Moeller"
date: "8/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# In lab 3, have them create their own markdown document. 

List of all the base R functions 
library(help = "base")
starting at abbreviate

# Research question
 - lm: researcher's Q was: what is the relationship between x and y?


# Object-oriented programming 
- Everything in R is an *object* 
- Every object has a *class* 
- The class of an object determines how code behaves on that object 

For example: `plot()` is an object with the class *function*. Functions do an action and create an output (in this case, a plot) when you put something inside the parentheses. If you run `plot()` on categorical data (class: factor), it will create a boxplot. If you run it on continuous data (class: numeric), it will create a scatterplot. 


# Finding duplicates 

# Finding outliers (data entry errors)

Do the data make sense? Look for values that *just can't be*. Functions that can be useful here are 

- for data frames: 
  - `summary()`
- for columns in a data frame: 
  - `range()`
  - `table()`
  - `unique()`
  
# Reordering rows in data 

To show rows in a sequential order, use `arrange()`

```{r}
penguin %>% 
  arrange(sample_number)
```


# 2.
Read in the messy penguin data ("penguins_raw.csv") and assign it to an object. 
```{r}

```

# 3. 
Assign a new object and rename all columns that have spaces or special characters.
```{r}

```

# 4. 
Using the object created in #3, investigate the values in the column `Sex`. Create a new object and remove any rows where the value in this column is not "Male" or "Female"
```{r}

```


# 5. 
Using the object created in #4, remove all rows that have an NA in any column and assign this as a new object. 
```{r}

```

# 6. 
Save your newly cleaned data as a CSV
```{r}

```