---
title: "Assignment 7 loops and functions"
author: "TBD"
date: "TBD"
output: html_document
---

# 1

Create a vector named `my_dat` with 6 non-consecutive values of fake data.
```{r}

```

# 2 

Write a "for" loop that multiplies each element of `my_dat` by 3 and store the results in a vector called `result` (hint: you will need to initialize `result` first). Call `result` to display your results. 
```{r}

```

# 3

Start with the vector provided. Write a "for" loop that overwrites the NAs with the one-half the previous value. Call the vector to display your results.
```{r}
vec <- c(18, rep(NA, 8))

```

# 4

Write a function that takes an input called `x`. Let the function add 1 to `x` then multiply that number by 2. 
```{r}

```

# 5

Write a function that takes inputs called `sd` and `n` that outputs `sd` divided by the square root of `n` (this is the calculation for standard error). Give your argument `n` a default value of 30. 
```{r}

```

# 6

Use your function from #5 to calculate the standard error of a sample with a standard deviation of 2.4 and a sample size of 18.
```{r}

```

# Final steps 

Make sure your name and date are correct in the header. Run spellcheck, then knit this document. Email the knitted HTML to paul.lukacs@umontana.edu.